Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-04-30 text/sections/00-abstract.tex
updated
Added new robustness check tables.

Public universities employ more lecturers and fewer professors than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows that the decline in state funding led to a substitution away from professors toward lecturers at US public universities.
Using a shift-share approach to instrument for state funding, I find that universities employ 4.4\% more lecturers per student following a 10\% funding cut.
This shift is accompanied by a reduction in assistant professors and full professors per student by 1.4\% and 1.2\%, respectively.
Incumbent professors' salaries, promotion rates, and quit rates at Illinois universities remain unaffected by funding cuts, indicating that the substitution arose from limiting the hiring of new tenure-track/tenured professors.
Stagnating state funding impacts public universities and faculty, likely contributing to the deterioration of student outcomes at public universities since the 1990s.

>>>>

Public universities employ more lecturers and fewer professors than at any other point in the last thirty years, relative to student enrolment.
At the same time, state funding for higher education has stagnated.
This paper shows that the decline in state funding led to a substitution away from professors toward lecturers at US public universities.
Using a shift-share approach to instrument for state funding, I find that universities employ 4.4\% more lecturers per student following a 10\% funding cut.
This shift is accompanied by a reduction in assistant and full professors by 1.4\% and 1.2\%  per student, respectively.
Incumbent professors' salaries, promotion rates, and quit rates at Illinois universities remain unaffected by funding cuts, indicating that the substitution arose from limiting the hiring of new tenure-track/tenured professors.
Stagnating state funding impacts public universities and faculty, likely contributing to deteriorating student outcomes at public universities seen since the 1990s.

### 2024-04-30 text/sections/05-results.tex
Added new robustness check tables.

Once the funding cut has passed, universities do not re-adjust to hiring more professors to make up for last year's substitution towards lecturers.
\autoref{fig:count-lp} shows local projection estimates, where the estimate is the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$ for later years $k = 0, \hdots, 10$.
These estimates correlespond to the elasticity estimates in \autoref{tab:facultycount-shock-reg} Panel B, where the outcome is faculty count per students in later years.
State funding cuts affectthe number of lecturers and professors at universities up to 9 years after the initial funding cut.
A 10\% state funding cut leads to 4.2\% (2.9--5.5 in a 95 percent confident interval) more lecturers per student the following year, and around 3.1\% (0.6--5.5 ``'') more lecturers nine years later.
A 10\% state funding cut leads to 1.5\% (0.6--2.2 in a 95 percent confident interval) fewer assistant professors per student the following year, and around 1.1\% (0.2--2.0 ``'') fewer nine years later;
the same cut leads to 1.2\% (0.6--1.8 in a 95 percent confident interval) fewer full professors per student the following year, and around 0.9\% (0.1--1.7 ``'') fewer nine years later.
The effects are a substitution with the composition of faculty, as the total number of faculty per students are relatively unaffected in the follow-on years.

@@

%\subsection{Robustness checks}
%\textbf{Describe robustness checks in one or two paragraphs here; see how Lovenheim$+$ did this section.}

@@

I take recent data on the total faculty hired at US public universities over 2011--2021 provided by \cite{wapman2022quantifying,wapman2022zenodo}.
\cite{wapman2022quantifying} study the composition of professor hiring in the years 2011--2021, and publicly released data on the number of professors (not lecturers) hired at US universities.\footnote{
    These data are not panel data, instead only give the total professors hired over the ten-year period 2011--2021.
}
I combine this measure, with total state funding for public universities across the same time period from IPEDS, to validate the claim that state funding cuts led to faculty substitution via limiting hiring of new professors.

>>>>

Once the funding cut has passed, universities do not re-adjust to hiring more professors to make up for last year's substitution towards lecturers.
\autoref{fig:count-lp} shows local projection estimates, where the estimate is the elasticity for professor count per student in year $t+k$ with respect to state funding in year $t$ for later years $k = 0, \hdots, 10$.
These estimates correspond to the elasticity estimates in \autoref{tab:facultycount-shock-reg} Panel B, where the outcome is faculty count per students in later years.
State funding cuts affect the number of lecturers and professors at universities up to 9 years after the initial funding cut.
A 10\% state funding cut leads to 4.2\% (2.9--5.5 in a 95 percent confident interval) more lecturers per student the following year, and around 3.1\% (0.6--5.5 confidence interval) more lecturers nine years later.
A 10\% state funding cut leads to 1.5\% (0.6--2.2 in a 95 percent confident interval) fewer assistant professors per student the following year, and around 1.1\% (0.2--2.0 confidence interval) fewer nine years later;
the same cut leads to 1.2\% (0.6--1.8 in a 95 percent confident interval) fewer full professors per student the following year, and around 0.9\% (0.1--1.7 confidence interval) fewer nine years later.
The effects are a substitution with the composition of faculty, as the total number of faculty per students are relatively unaffected in the follow-on years.

\textbf{Paragraph comparing to any previous estimates, as do Lovenheim 2020 page 22.}

@@

\textbf{Paragraph comparing to any previous estimates, as do Lovenheim 2020 page 22.}

\subsection{Robustness Checks}
\label{sec:results-robustness}
\textbf{Describe robustness checks in one or two paragraphs here; see how Lovenheim$+$ did this section.}

These tables describe the same analysis, with other controls added in, and with State $+$ Year fixed effects --- instead of Institution $+$ Year fixed effects.
\autoref{tab:firststage-robustness-checks},
\autoref{tab:facultycount-robustness-checks},
\autoref{tab:facultycount-rawcount-robustness-checks}.
Uses the FTE enrolment measure, with measures of selectivity.

\autoref{tab:facultycount-shock-private-robustness} describes the results of using private universities in a falsification test, by assigning their base share as the average of similarly selective public universities in their state, and using this as an instrument for total revenues (as state funding is uncommon for private universities).
Note, the first-stage is not strong, and the IV estimates are all (encouragingly) not significant at any standard frequentist cut-off.

\autoref{tab:facultycount-heterogeneity} shows  the effects of state funding cuts, separated among universities with different selectivity rankings.
The most selective (rank 1-2, including Berkeley to SUNY Binghamton) have no faculty substitution effects, but even among selective universities (rank 3-4, including Clemson University to University of Dallas) there are strong substitution effects.
1 indicates “most competitive,” 2 is “highly competitive plus,” 3 is “highly competitive,” and 4 is “very competitive plus.”

@@

I take recent data on the total faculty hired at US public universities over 2011--2021 provided by \cite{wapman2022quantifying};
they studied the composition of professor hiring in the years 2011--2021, and publicly released data on the number of professors (not lecturers) hired at US universities \citep{wapman2022zenodo}.\footnote{
    These data are not panel data, instead only give the total professors hired over the ten-year period 2011--2021.
}
I combine this measure, with total state funding for public universities across the same time period from IPEDS, to validate the claim that state funding cuts led to faculty substitution via limiting hiring of new professors.

### 2024-04-30 text/sections/07-conclusion.tex
Added new robustness check tables.

This work contributes to the literature along two primary dimensions.
First, this work provides an explanation for the increased reliance on lecturers and away from tenure-track and tenured professors at US public universities, by isolating the effects funding cuts from state-wide funding shifts in both the short-- and long-run.
Second, this work investigates how individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes, in the Illinois university system, and show that incumbent faculty were relatively unaffected by state funding cuts to their employers.

>>>>

This work contributes to the literature along two primary dimensions.
First, this work provides an explanation for the increased reliance on lecturers and away from tenure-track and tenured professors at US public universities, by isolating the effects funding cuts from state-wide funding shifts in both the short-- and long--run.
Second, this work investigates how individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes, in the Illinois university system, and show that incumbent faculty were relatively unaffected by state funding cuts to their employers.

### 2024-04-30 text/sections/09-figures-tables.tex
Added new robustness check tables.

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Effects of Changes in State Funding on University Faculty Composition, IPEDS 1990--2021, OLS and 2SLS Estimates.}

@@

    \label{tab:facultycount-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on faculty outcomes, using the state funding shift-share to instrument for state funding in the columns labelled 2SLS.
    Each observation is a public university-year, in the IPEDS data.
    Panel A shows the effect of a fall in state funding \$$-1,000$ per student on the number of professors --- i.e.,
    an extra \$$1,000$ per student leads to 6 fewer lecturers according to column 2.
    Panel B shows the effect of a $10$\% change in state funding per student at the university on the 10\% change in the number of professors per students --- i.e.,
    an extra 10\% of state funding per student leads to 4.37\% fewer lecturers per student according to column 2.
    Outcome-mean is the mean of the outcome, for Panel A the number of professors per student, for Panel B the number of faculty per student.
    Panel B uses $\log$ faculty count per student as the outcome, though the outcome mean is count of faculty per student (not in $\log$ terms).
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included through--out.
\end{table}

@@

    \justify
    \footnotesize
    \textbf{Note}: 
    These tables show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on faculty outcomes at Illinois universities, using the state funding shift-share to instrument for state funding.
    The shift-share instrument is based in the year the professor was hired, following definition in \autoref{sec:iv-model-indiv}.
    Each observation is a faculty member--year at an Illinois public university, where funding data come from IPEDS and faculty outcomes from IBHED data.
    The panels shows the effect of a $1$\% change in state funding per student at the university on faculties' salaries (base salary $+$ benefits), promotion rate (e.g., assistant professor to associate professor), and rate of leaving their university (i.e., by quitting, retiring, or moving to another university).
    Standard errors are clustered at the university-year of hire level, and university $+$ year of hire fixed effects are included through--out.
\end{table}

>>>>

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Effects of State Funding on Faculty Counts, IPEDS 1990--2021, OLS and 2SLS Estimates.}

@@

    \label{tab:facultycount-shock-reg}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on faculty outcomes, using the state funding shift-share to instrument for state funding in the columns labelled 2SLS.
    Each observation is a public university-year, in the IPEDS data.
    Panel A shows the effect of a fall in state funding \$$-1,000$ per student on the number of professors --- i.e.,
    an extra \$$1,000$ per student leads to 6 fewer lecturers according to column 2.
    Panel B shows the effect of a $10$\% change in state funding per student at the university on the 10\% change in the number of professors per students --- i.e.,
    an extra 10\% of state funding per student leads to 4.37\% fewer lecturers per student according to column 2.
    Outcome-mean is the mean of the outcome, for Panel A the number of professors per student, for Panel B the number of faculty per student.
    Panel B uses $\log$ faculty count per student as the outcome, though the outcome mean is count of faculty per student (not in $\log$ terms).
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included through-out.
\end{table}

@@

    \justify
    \footnotesize
    \textbf{Note}: 
    These tables show the second stage 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on faculty outcomes at Illinois universities, using the state funding shift-share to instrument for state funding.
    The shift-share instrument is based in the year the professor was hired, following definition in \autoref{sec:iv-model-indiv}.
    Each observation is a faculty member--year at an Illinois public university, where funding data come from IPEDS and faculty outcomes from IBHED data.
    The panels shows the effect of a $1$\% change in state funding per student at the university on faculties' salaries (base salary $+$ benefits), promotion rate (e.g., assistant professor to associate professor), and rate of leaving their university (i.e., by quitting, retiring, or moving to another university).
    Standard errors are clustered at the university-year of hire level, and university $+$ year of hire fixed effects are included through-out.
\end{table}

### 2024-04-30 text/sections/10-appendix.tex
Added new robustness check tables.

This approach leverages an insight, made available by level of the data: that an individual professor is affected by changes in university revenues after they have joined the university.
\autoref{sec:iv-model-uni} considers the number of professors employed by the university; whether a professor becomes employed at the university is likely affected by that university's finances.
The formulation here takes as given that the professor is already employed at the university, and then projects the effect of changes in state funding on these \textit{incumbent} professors following the state funding shift-share.
\autoref{tab:firststage-illinois} presents the first-stage results in Illinois data .

@@

>>>>

This approach leverages an insight, made available by level of the data: that an individual professor is affected by changes in university revenues after they have joined the university.
\autoref{sec:iv-model-uni} considers the number of professors employed by the university; whether a professor becomes employed at the university is likely affected by that university's finances.
The formulation here takes as given that the professor is already employed at the university, and then projects the effect of changes in state funding on these \textit{incumbent} professors following the state funding shift-share.
\autoref{tab:firststage-illinois} presents the first-stage results in Illinois data.

@@

\newpage
\subsection{IPEDS Robustness Checks}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Effects of State Funding on Faculty Counts, IPEDS 1990--2021, IV Estimates by Institution Selectivity.}

    \makebox[\textwidth][c]{
        \begin{tabular}{@{\extracolsep{5pt}}lcccccl} 
            \\[-1.8ex]\hline 
            \hline \\[-1.8ex]
            & First-stage & Lecturers & Asst. Professors & Full Professors & All Faculty & Observations \\ 
            \cline{2-7} 
            \\[-1.8ex]
            \input{tables/facultycount-rawcounts-heterogeneity.tex}
            \\[-1.8ex] \hline 
            \hline 
        \end{tabular}}

        \vspace{0.2cm}
    \textbf{Panel B: units in log \$ per student}

    \makebox[\textwidth][c]{
        \begin{tabular}{@{\extracolsep{5pt}}lcccccl} 
            \\[-1.8ex]\hline 
            \hline \\[-1.8ex]
            & First-stage & Lecturers & Asst. Professors & Full Professors & All Faculty & Observations \\ 
            \cline{2-7} 
            \\[-1.8ex]
            \input{tables/facultycount-heterogeneity.tex}
            \\[-1.8ex] \hline 
            \hline 
        \end{tabular}}
    \label{tab:facultycount-heterogeneity}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the IV estimates of regression specification \eqref{eqn:secondstage}, 
    in the same manner as \autoref{tab:facultycount-shock-reg}, but restricting to institutions ranked as most selective, selective, and unranked by \cite{barrons2009}.
    For example, Row 1 of panel B shows that a $10\%$ cut in the state shift-share leads to a fall of $9.44\%$ in state funding for the public universities ranked as ``most selective.''
\end{table}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{First-stage Robustness Checks for Effects of State Funding Shift-Share on State Funding, OLS Estimates.}
    \makebox[\textwidth][c]{\input{tables/firststage-robustness-checks.tex}}
    \label{tab:firststage-robustness-checks}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:firststage}, showing the effect of state funding shift-share on each university's state funding.
    This table differs from the main analysis by replacing University $+$ Year fixed effects with State $+$ Year fixed effects, measuring enrolment in full-time equivalent (FTE), and including controls for (1)
    State Funding as a percent of total funding in base period, (2) acceptance rate in mid-2000s, (3) 6--year completion rate in mid-2000s, (4) tuition revenue, (5) enrolment measured by full-time equivalent FTE, (6) total public university enrolment in the entire state, (7) percent of public university enrolment for the state enrolled at this university.
    The first column uses the raw count specification, and the second column uses the $\log$ specification for percentage terms.
\end{table}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Robustness Checks for Effects of State Funding Cuts on Faculty Composition, OLS and 2SLS Estimates in $\log$ Units.}
    \makebox[\textwidth][c]{
        %\small
        \input{tables/facultycount-robustness-checks.tex}}
    \label{tab:facultycount-robustness-checks}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on number of faculty per student in Illinois universities, using the funding shift-share to instrument for state funding in the columns labelled 2SLS.
    This table differs from the main analysis by replacing University $+$ Year fixed effects with State $+$ Year fixed effects, measuring enrolment in full-time equivalent (FTE), and including controls for (1)
    State Funding as a percent of total funding in base period, (2) acceptance rate in mid-2000s, (3) 6--year completion rate in mid-2000s, (4) tuition revenue, (5) enrolment measured by full-time equivalent FTE, (6) total public university enrolment in the entire state, (7) percent of public university enrolment for the state enrolled at this university.
\end{table}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Robustness Checks for Effects of State Funding Cuts on Faculty Composition, OLS and 2SLS Estimates in Raw Count Units.}
    \makebox[\textwidth][c]{
        \small
        \input{tables/facultycount-rawcount-robustness-checks.tex}}
    \label{tab:facultycount-rawcount-robustness-checks}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, showing the effect of state funding changes on count of faculty in Illinois universities, using the funding shift-share to instrument for state funding in the columns labelled 2SLS.
    This table differs from the main analysis by replacing University $+$ Year fixed effects with State $+$ Year fixed effects, measuring enrolment in full-time equivalent (FTE), and including controls for (1)
    State Funding as a percent of total funding in base period, (2) acceptance rate in mid-2000s, (3) 6--year completion rate in mid-2000s, (4) tuition revenue, (5) enrolment measured by full-time equivalent FTE, (6) total public university enrolment in the entire state, (7) percent of public university enrolment for the state enrolled at this university.
\end{table}

\begin{table}[H]
    \singlespacing
    \centering
    \caption{Effects of State--Wide Funding Changes on Private University Faculty Counts, IPEDS 1990--2021, OLS and 2SLS Estimates.}

    \makebox[\textwidth][c]{\input{tables/facultycount-shock-private-robustness-rawcount.tex}}

    \makebox[\textwidth][c]{\input{tables/facultycount-shock-private-robustness.tex}}

    \label{tab:facultycount-shock-private-robustness}
    \justify
    \footnotesize
    \textbf{Note}:
    These tables show the second stage OLS and 2SLS estimates of regression specification \eqref{eqn:secondstage}, among private universities --- as described in \autoref{sec:results-robustness}.
    Each observation is a public university-year, in the IPEDS data.
    Panel A shows the effect of a fall in state funding \$$-1,000$ per student on the number of professors --- i.e.,
    an extra \$$1,000$ per student leads to 6 fewer lecturers according to column 2.
    Panel B shows the effect of a $10$\% change in state funding per student at the university on the 10\% change in the number of professors per students --- i.e.,
    an extra 10\% of state funding per student leads to 4.37\% fewer lecturers per student according to column 2.
    Outcome-mean is the mean of the outcome, for Panel A the number of professors per student, for Panel B the number of faculty per student.
    Panel B uses $\log$ faculty count per student as the outcome, though the outcome mean is count of faculty per student (not in $\log$ terms).
    Standard errors are clustered at the state-year level, and university $+$ year fixed effects are included through--out.
\end{table}

### 2024-04-30 text/tables/firststage-balance-count.tex
Adjustments.
Wiritng updates, new title.
Added new robustness check tables.

\begin{tabular}{@{\extracolsep{5pt}}lcccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{8}{c}{Dependent Variables: University Characteristics} \\ 
\cline{2-9}
\\[-1.8ex] & Enrolment & State & Total & Non-inst. & Lecturers & Assistant & Full & All \\ 
\\[-1.8ex] &           & Funding & Revenues & Revenues & Lecturers & Professors & Professors & Professors \\
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8)\\ 
\hline \\[-1.8ex] 
 Funding & 0.517   & $-$9,217 & $-$7,633 & 886 & 0.007 & $-$0.001 & $-$0.003 & 0.004 \\ 
 Shock   & (0.199) & (2,452) & (12,771) & (4,093) & (0.002) & (0.001) & (0.003) & (0.005) \\ 
 \hline \\[-1.8ex] 
Uni. + Year \\
fixed effects? & Yes & Yes & Yes & Yes & Yes & Yes & Yes & Yes \\
Observations & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 \\ 
R$^{2}$ & 0.960 & 0.950 & 0.881 & 0.932 & 0.745 & 0.886 & 0.973 & 0.954 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular}

>>>>

\begin{tabular}{@{\extracolsep{5pt}}lcccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{8}{c}{Dependent Variables: University Characteristics} \\ 
\cline{2-9} 
\\[-1.8ex] & Enrolment & State Funding & Total revenues & Non-inst. revenues & Lecturers & Assistant professors & Full professors & All professors \\ 
\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8)\\ 
\hline \\[-1.8ex] 
 State Funding & 0.517 & $-$9,217.091 & $-$7,633.377 & 1,818.090 & 0.007 & $-$0.001 & $-$0.003 & 0.004 \\ 
  & (0.199) & (2,451.719) & (12,771.130) & (4,774.483) & (0.002) & (0.001) & (0.003) & (0.005) \\ 
 \hline \\[-1.8ex] 
Uni. + Year fixed effects? & Yes & Yes & Yes & Yes & Yes & Yes & Yes & Yes \\ 
Observations & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 & 17,012 \\ 
R$^{2}$ & 0.960 & 0.950 & 0.881 & 0.922 & 0.745 & 0.886 & 0.973 & 0.954 \\ 
\hline 
\hline \\[-1.8ex] 
\end{tabular}
